---
title: "Week 1"
author: "Tad Dallas"
includes:
  in_header:
    - \usepackage{lmodern}
output:
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    toc: yes
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: tango
    theme: journal
---



 
((use this lecture to discuss the concepts of stable and unstable equilibria))


## Exponential growth

```{r}

n0 <- 1:100
r <- 1


expoGrowth <- function(n, r){
  n+(n*r)  
}



#effect of growth rate
plot(n0, expoGrowth(n0, 1.15),
  type='l', las=1, 
  xlab='Population size at time t', 
  ylab='Population size at time t+1', 
  col='dodgerblue')
lines(n0, expoGrowth(n0,1),
  col='firebrick')
lines(n0, expoGrowth(n0,0.75),
  col='green')

legend('bottomright', paste(expression(lambda), c('1.15', '=1', '=0.75')),
  pch=16, col=c('dodgerblue', 'firebrick', 'green'))
 

```






```{r}

expoDynamics <- function(n,r, steps=100){
  ret <- c()
  ret[1] <- n
  for(i in 1:steps){
    ret[i+1] <- expoGrowth(ret[i], r)
  }
  return(ret)
}

```



```{r}

t <- 50
n0 <- 20

#effect of growth rate
plot(1:(t+1), expoDynamics(n0, 1.15, steps=t),
  type='l', las=1, 
  xlab='Time', 
  ylab='Population size', 
  col='dodgerblue')
lines(1:(t+1), expoDynamics(n0, 1, steps=t),
  col='firebrick')
lines(1:(t+1), expoDynamics(n0, 0, steps=t),
  col='green')

legend('topleft', bty='n', 
  paste('r = ', c('1.15', '1', '0')),
  pch=16, col=c('dodgerblue', 'firebrick', 'green'))


```

















## Logistic growth

```{r}

n0 <- 1:150
k <- 20
r <- 1


logisticGrowth <- function(n, r, k){
  n*exp(r*(1-(n / k)))
}


colz <- c(grey(0.1,0.9), 'dodgerblue', 'firebrick', 'forestgreen')
#effect of growth rate
plot(n0, logisticGrowth(n0,1.15,20),
  type='l', las=1, 
  xlab='Population size at time t', 
  ylab='Population size at time t+1', 
  col=colz[1])
lines(n0, logisticGrowth(n0,1,20),
  col=colz[2])
lines(n0, logisticGrowth(n0,0.75,20),
  col=colz[3])
legend('topright', bty='n',
  c('r=0.75', 'r=1', 'r=1.15'),
  pch=16, col=colz[c(3,2,1)])





#effect of carrying capacity
plot(n0, logisticGrowth(n0,1, 50),
  type='l', las=1, ylim=c(0,100),
  xlab='Population size at time t', 
  ylab='Population size at time t+1', 
  col=colz[2])
lines(n0, logisticGrowth(n0,1,100),
  col=colz[1])
lines(n0, logisticGrowth(n0,1,10),
  col=colz[3])
legend('topright', bty='n',
  c('k=10', 'k=50', 'k=100'),
  pch=16, col=colz[c(3,2,1)])




#Look at the peaks of the growth (where is the maximum population size here?)

plotSegs <- function(kx,ky, color){
  segments(x0=0,x1=kx, y0=ky,y1=ky, col=color, lwd=2, lty=3)
  segments(x0=kx,x1=kx, y0=0,y1=ky, col=color, lwd=2, lty=3)
  points(kx,ky, pch=16, cex=2, col=color)
}

plotSegs(10,10, color=colz[3])
plotSegs(50,50, color=colz[2])
plotSegs(100,100, color=colz[1])



# But why is this? What situations would cause this to not be the case?

plot(n0, logisticGrowth(n0,1.5, 50),
  type='l', las=1, ylim=c(0,150),
  xlab='Population size at time t', 
  ylab='Population size at time t+1', 
  col=colz[2])
lines(n0, logisticGrowth(n0,1.5,100),
  col=colz[1])
lines(n0, logisticGrowth(n0,1.5,10),
  col=colz[3])
legend('topright', bty='n',
  c('k=10', 'k=50', 'k=100'),
  pch=16, col=colz[c(3,2,1)])



#Look at the peaks of the growth (where is the maximum population size here?)
plotSegs(10,10, color=colz[3])
plotSegs(50,50, color=colz[2])
plotSegs(100,100, color=colz[1])



# this line intersects points where population change from t to t+1 is 0. These are equilibria.
abline(a=0,b=1)

```






Alright. So now we can look at the actual dynamics across many generations.


```{r}

logisticDynamics <- function(n,r,k, steps=100){
  ret <- c()
  ret[1] <- n
  if(length(r) == 1){
    r <- rep(r, steps)
  }
  for(i in 1:(steps-1)){
    ret[i+1] <- logisticGrowth(ret[i], r[i], k)
  }
  return(ret)
}




```


```{r}

stps <- 20

plot(1:stps,
 logisticDynamics(n=30, r=1, k=30, steps=stps),
  type='l', las=1, ylim=c(0,50),
  xlab='Time', 
  ylab='Population size', 
  col=1)

#sapply(seq(1,25,by=1), function(x){
# lines(logisticDynamics(n=x, r=1, k=30, steps=stps), col='firebrick')
#})


sapply(seq(5,30,by=1), function(x){
 lines(logisticDynamics(n=x, r=1, k=30, steps=stps), col='dodgerblue')
})

sapply(seq(5,30,by=1), function(x){
 lines(logisticDynamics(n=x, r=rep(c(0.5,1.5),5), k=30, steps=stps), col='firebrick')
})



```






What if growth rate is not 1?

r=0


```{r}

stps <- 20

plot(1:stps,
 logisticDynamics(n=30, r=0, k=30, steps=stps),
  type='l', las=1,ylim=c(0,100),
  xlab='Time', 
  ylab='Population size', 
  col=1)

sapply(seq(1,25,by=1), function(x){
 lines(logisticDynamics(n=x, r=0, k=30, steps=stps), col='firebrick')
})

sapply(seq(35,100,by=5), function(x){
 lines(logisticDynamics(n=x, r=0, k=30, steps=stps), col='dodgerblue')
})

```







r=1.5

```{r}

stps <- 20

plot(1:stps,
 logisticDynamics(n=29, r=1.5, k=30, steps=stps),
  type='l', las=1,ylim=c(0,100),
  xlab='Time', 
  ylab='Population size', 
  col=1)

sapply(seq(1,25,by=1), function(x){
 lines(logisticDynamics(n=x, r=1.5, k=30, steps=stps), col='firebrick')
})

sapply(seq(35,100,by=5), function(x){
 lines(logisticDynamics(n=x, r=1.5, k=30, steps=stps), col='dodgerblue')
})


```







r=2

```{r}

stps <- 50

plot(1:stps,
 logisticDynamics(n=30, r=2, k=30, steps=stps),
  type='l', las=1,ylim=c(0,100),
  xlab='Time', 
  ylab='Population size', 
  col=1)

plot(1:stps,
 logisticDynamics(n=29, r=2, k=30, steps=stps),
  type='l', las=1,ylim=c(0,100),
  xlab='Time', 
  ylab='Population size', 
  col=1)

sapply(seq(1,25,by=1), function(x){
 lines(logisticDynamics(n=x, r=2, k=30, steps=stps), col='firebrick')
})

sapply(seq(35,100,by=5), function(x){
 lines(logisticDynamics(n=x, r=2, k=30, steps=stps), col='dodgerblue')
})


```





(between 3 and 3.449 -- oscillates between 2 values)

r=3.2

```{r}

stps <- 20

plot(1:stps,
 logisticDynamics(n=20, r=3.2, k=30, steps=stps),
  type='l', las=1,ylim=c(0,100),
  xlab='Time', 
  ylab='Population size', 
  col=1)


sapply(seq(1,100,by=5), function(x){
 lines(logisticDynamics(n=x, r=3.2, k=30, steps=stps), col='firebrick')
})

```









(onset of chaos)
r > 3.56

```{r}

stps <- 100

plot(1:stps,
 logisticDynamics(n=30, r=4.75, k=30, steps=stps),
  type='l', las=1,ylim=c(0,150),
  xlab='Time', 
  ylab='Population size', 
  col=1)


plot(1:stps,
 logisticDynamics(n=29, r=4.75, k=30, steps=stps),
  type='l', las=1,ylim=c(0,150),
  xlab='Time', 
  ylab='Population size', 
  col=1)

```






































## Matrix model

((give this as an example problem to let the students work through it))

First, build a simple projection matrix


```{r}

projMatrix <- matrix(
  c(
    0.2,     1,    0.25,
    0.3,     0,    0,
    0,       0.3,  0.6
  )
  ,nrow=3,ncol=3,byrow=T
)


abund0 <- matrix(c(20,20,20), ncol=1)

```





Simulate one generation into the future.

```{r}

abund1 <- projMatrix  %*% abund0 


```



Simulate one more generation 

```{r}

projMatrix %*% abund1


```





Simulate many generations to examine dynamics 

```{r}

getStageDynamics <- function(projMatrix, abund, steps=100){
  ret <- matrix(0, ncol=3, nrow=steps+1)
  ret[1,] <- abund
  for(i in 1:steps){
    ret[i+1, ] <- projMatrix %*% matrix(ret[i,],ncol=1) 
  }
  return(ret)
}

```




```{r}

stageDynamics <- getStageDynamics(projMatrix, abund0, steps=50)


plot(stageDynamics[,1], type='l', lwd=2,
  col=colz[1],
  ylab='Abundance', xlab='Time')
lines(stageDynamics[,2], lwd=2, col=colz[2])
lines(stageDynamics[,3], lwd=2, col=colz[3])
legend('topright', pch=16, col=colz[1:3],
  c('Young', 'Middle', 'Old'), bty='n')

```












Explore some of the dynamics with the projection matrix and equilibrium points. 


```{r}





```

