---
title: "Week 4"
author: "Tad Dallas"
includes:
  in_header:
    - \usepackage{lmodern}
output:
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    toc: yes
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: tango
    theme: journal
---



 




# How do we model the niche and corresponding geographic distribution of a species?

1) take occurence points
2) extract environmental layers
3) examine the niche in reduced (or not) dimensions
4) train regression model on species occurrences (discuss presence-absence versus presence-background data)
5) predict with changing environment or in new area









1) take occurence points

```{r}





```








2) extract environmental layers

```{r}





```







3) examine the niche in reduced (or not) dimensions

```{r}





```








4) train regression model on species occurrences (discuss presence-absence versus presence-background data)

```{r}





```










5) predict with changing environment or in new area

```{r}





```










## TILMAN R*

library(ecostudy)
## One resource
mod <- tilman(S = 20, a = 0.5, r = c(2,1), m = 0.5, k = c(5,0.5), c = 0.1)
mod
plot(mod)
## Two resources
mod2 <- update(mod, S=c(20,20), k=c(5,0.5, 1,2), c = c(0.15, 0.1, 0.1, 0.2))
tilmandiagr(mod2)
tilmandiagr(mod2, S = c(10,20), add = TRUE, time = 100)
tilmandiagr(mod2, S = c(20,10), add = TRUE, time = 100)
op <- par(mfrow=c(3,1), mar=c(4,4,0,1)+.1)
plot(mod2)
plot(mod2, S = c(10,20))
plot(mod2, S = c(20,10))
par(op)








## LOTKA-VOLTERRA COMPETITION

mod <- lotkacomp(0.8, 0.6, 20, 20)
plot(mod)

## Add trajectory line and starting point
lines(mod, 1, 1, col="green", lwd=2)
points(1, 1, pch=16, col="green")
## Plot populations against time
plot(traj(mod, 1, 1))
## Textual output
mod
summary(mod)








