---
title: "Week 4"
author: "Tad Dallas"
includes:
  in_header:
    - \usepackage{lmodern}
output:
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    toc: yes
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: tango
    theme: journal
---



 




# How do we model the niche and corresponding geographic distribution of a species?

1) take occurence points
2) extract environmental layers
3) examine the niche in reduced (or not) dimensions
4) train regression model on species occurrences (discuss presence-absence versus presence-background data)
5) predict with changing environment or in new area









1) take occurence points

```{r}





```








2) extract environmental layers

```{r}





```







3) examine the niche in reduced (or not) dimensions

```{r}





```








4) train regression model on species occurrences (discuss presence-absence versus presence-background data)

```{r}





```










5) predict with changing environment or in new area

```{r}





```



